---
title: "Annotation Routine"
output: github_document
vignette: >
  %\VignetteIndexEntry{Routine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "vignettes/figures/",
  out.width = "100%"
)
```

# A routine for developing annotation instructions

In this vignette I am presenting a general routine for developing intersubjectively valid annotation instructions for your llm-based annotation.
For these steps I will use the example data from another project to illustrate steps 1 - 3. I will note peculiarities in the dataset (or where I stray from this routine) where necessary.

## Step 1: Write a first draft of annotation instructions

Write annotation instructions. They work similar to codebooks for human annotation: They explain the purpose of your measurement, what data to expect, how concepts are defined and how they are operationalized.

Unlike a human codebook, bacchuss is structured for measuring one dimension at a time. If you have multiple dimensions of interest, I suggest annotating these dimensions in separate steps with separate annotation instructions. While it is possible to have a multi-dimensional output (and I might implement it at a later date), keeping annotation instructions separate has the following advantages:

- We safe space within the model's context window for each call
- We avoid coding instructions of an earlier step confusing later annotation.

For Codebook structure, I suggest the following general structure:

1. Role description: a neutral description of the general task
2. Context: What data the LLM can expect - do you add metadata? What should it focus on? What is the data source? Often it is helpful to instruct the model to be literal and search for explicit expressions of the concept and not interpret the text using context knowledge.
3. An explanation for each label: A general definition, distinguishing criteria (with different ways the concept is expressed if necessary), typical markers/keywords
4. Decision rules: Often, you will identify cases where you need clear demarkation rules to distinguish labels. Decision rules are the place for explaining those decision rules
5. Output format: bacchuss supports 2 output formats: Either it uses regular expressions, then it expects the format "Label: <Label> \n Explanation: <Explanation>", or it expects JSON output with "label" and "explanation" as separate fields. We will use the former format here - for JSON mode see documentation. I will add a JSON vignette in a later version.

Expect to have a general idea of 1, 2, 3 and 5 before moving on to the next step - adjust all of them and add 4 gradually as you see more hard cases.

I usually send a short reminder of coding instructions after the text to code. This should be a one- or two-sentence summary of the most important points of your codebook.

For a general overview how to write good annotation instructions I suggest the literature on human annotation - for example, Kimberly Neuendorf's "The Content Analysis Guidebook".
If there are previous operationalizations for your concept of interest, you can translate them to annotation instructions by putting them into the format above. You will probably still need to do the next steps and improve the instructions.
For text data you are unfamiliar with, it makes sense to take a qualitative survey of how different expressions of your dimension of interest occur in your texts before writing a codebook from scratch - the context of your data can change how concepts are expressed within the text.

## Step 2.1: Test draft annotation instructions

When you have a first draft for annotation instructions, run annotation on a small sub-sample of your data. 

```{r example2_1, eval = FALSE}
library(bacchuss)

annotation_test_2_1 <- bacchuss(example_set, input_column = "paragraphs",
                        instructions = example_instructions,
                        reminder = example_reminder,
                        model = "mistral-nemo",
                        host = NULL,
                        api_key = NULL)

View(example_zero)
```

Visually inspect your results:

- which cases were correctly labelled?
- are there clear cases not described in your coding instructions?
- Note down general problems of your instructions
- Improve annotation instructions based on those notes - add decision rules, clarify your concept

(Note: In this vignette I am here using the same data I will later use for step 3. In your own projects, instead use a random sample from your data here.)

### Note: Repairing mistyped labels:

As noted in the readme, small LLMs have a tendency to mistype labels due to token drift. Even if you tell the model to annotate two labels, it might output a third, based on jumps of logic ("I am told to label fear and admiration, here is anger, so I code anger") or simple misspelling ("amdiration"). Use the count function in dplyr to get a count of all labes in your output and recode false labels (for example, to turn "feer" into "fear"). 
Rerun annotation for cases with no clear label.
If more than 5 percent of your output is false labels, try adjusting the reminder or the instructions or both. I will later at a vignette how to do this.

## Step 2.2: Check for hard cases

Once you have improved your annotation instructions in step 2.1, it is time to test whether there are hard cases the llm struggles with - cases where it would output different labels because the annotation instructions give some leeway for interpretation. In human annotation, we often note down such hard cases during codebook improvement - this step is meant to do the same for LLM annotation.
We run annotation with higher temperature and multiple runs and note disagreement. 

The function briseus (another name for the god Bacchus, translating to "he who prevails") - is meant for this purpose.

```{r example2_2, eval = FALSE}
library(bacchuss)

annotation_test_2_2 <- briseus(example_set, input_column = "paragraphs", n_runs = 5,
                        temperature = 0.7,
                        instructions = example_instructions,
                        reminder = example_reminder,
                        model = "mistral-nemo",
                        host = NULL,
                        api_key = NULL)

View(example_zero)
```

The function used here will give you the final label it arrived on, labels for all runs, and agreement. Sort by agreement and inspect the cases that have low agreement between runs.

Visually inspect:
- Why are those cases hard for annotation? Can the rules be clarified?
- which cases were correctly labelled?
- are there clear cases not described in your coding instructions?
- Note down general problems of your instructions
- Improve annotation instructions based on those notes - add decision rules, clarify your concept

(Note: In this vignette I am here using the same data I will later use for step 3. In your own projects, instead use a random sample from your data here. It is probably best to draw new samples between each step to cover a broader range of examples.)

## Repeat steps 2.1 and 2.2 until your annotation instructions cover the concept broadly

This does not mean that reliability is high at this point - you may need to add chain of thought prompting or few-shot examples to reach acceptable reliability. Note how the final annotation instructions of the sample data code almost none of the fear cases - that only really started working with few-shot examples.

Instead, look

## Step 3: Test non-representative sample of hard and soft cases

Once you think you have broadly covered your concept and it's the time for refining your instructions, we move on to step 3.
For this step, you collect sample cases from rounds 2.1 + 2.2, decide on correct labels (in this step, you can use consensus coding - all project members discuss and agree on which labels should be correct for this small set), and use this dataset to further optimization.

For these steps, I have added a sample dataset of human curated data from a previous project.

## Step 3.1: Test improvements of the annotation instructions

Run bacchuss with your improved annotation instructions and measure how well it conforms to your human standard labels.

```{r example3_1, eval = FALSE}
library(bacchuss)

example_zero <- bacchuss(example_set, input_column = "paragraphs",
                        instructions = example_instructions,
                        reminder = example_reminder,
                        model = "mistral-nemo",
                        host = NULL,
                        api_key = NULL)

View(example_zero)
```

Then compare your data with the human standard labels:

```{r example3_1_2, eval = FALSE}
example_zero$fear_human <- example_zero$Emotion == "Angst/Furcht (1)"

example_zero$fear_llm <- example_zero$labels == "Angst/Furcht (1)"

example_zero$adm_human <- example_zero$Emotion == "Bewunderung/tiefer Respekt (2)"

example_zero$adm_llm <- example_zero$labels == "Bewunderung/tiefer Respekt (2)"

mt_fear_3_1 <- table(example_zero$fear_llm, example_zero$fear_human)
caret::confusionMatrix(mt_fear_3_1, mode = "prec_recall", positive = "TRUE")

mt_admiration_3_1 <- table(example_zero$adm_llm, example_zero$adm_human)
caret::confusionMatrix(mt_admiration_3_1, mode = "prec_recall", positive = "TRUE")

```

You will see that with only the annotation instructions, the rules for detecting fear seem to be very unreliable - we barely pick up fear.

(In the example data, we were only interested in the categories 1 and 2, and treated 4 and 9 as "other" in our analysis. That's why I recode the data in this step to give separate results for only those two labels. For your own data, it usually makes sense to report at least separate scores for each label.)

## Step 3.2: Add Chain-of-Thought, few-shot examples if necessary

You can add instructions to output an Explanation for why a label was chosen - a form of chain-of-thought prompting. Instruct the model to not just output a label, but also an explanation, and set explanation to true.

```{r example3_2_1, eval = FALSE}
library(bacchuss)

example_zero_cot <- bacchuss(example_set, input_column = "paragraphs",
                            instructions = example_instructions_expl,
                            reminder = example_reminder_expl,
                            explanations = TRUE,
                            model = "mistral-nemo",
                            host = NULL,
                            api_key = NULL)

View(example_zero_cot)
```

This often improves your results. You can also add few-shot examples:

```{r example3_2_2, eval = FALSE}
library(bacchuss)

example_few <- bacchuss(example_set, input_column = "paragraphs",
                        instructions = example_instructions,
                        examples = example_fewshot$input,
                        codes = example_fewshot$label,
                        reminder = example_reminder, 
                        model = "mistral-nemo",
                        host = NULL,
                        api_key = NULL)

View(example_few)
```

And you can combine few-shot and chain-of-thought by providing explanations for each few-shot label (and using annotation instructions that instruct to output explanations).

```{r example3_2_3, eval = FALSE}
library(bacchuss)

example_few_cot <- bacchuss(example_set, input_column = "paragraphs",
                            instructions = example_instructions_expl,
                            examples = example_fewshot$input,
                            explanations = example_fewshot$explanation,
                            codes = example_fewshot$label,
                            reminder = example_reminder_expl,
                            model = "mistral-nemo",
                            host = NULL,
                            api_key = NULL)

View(example_few_cot)
```

As you will see, this improves validity:

```{r example3_2_3_2, eval = FALSE}
example_few_cot$fear_human <- example_few_cot$Emotion == "Angst/Furcht (1)"

example_few_cot$fear_llm <- example_few_cot$labels == "Angst/Furcht (1)"

example_few_cot$adm_human <- example_few_cot$Emotion == "Bewunderung/tiefer Respekt (2)"

example_few_cot$adm_llm <- example_few_cot$labels == "Bewunderung/tiefer Respekt (2)"

mt_fear_few_cot <- table(example_few_cot$fear_llm, example_few_cot$fear_human)
caret::confusionMatrix(mt_fear_few_cot, mode = "prec_recall", positive = "TRUE")

mt_admiration_few_cot <- table(example_few_cot$adm_llm, example_few_cot$adm_human)
caret::confusionMatrix(mt_admiration_few_cot, mode = "prec_recall", positive = "TRUE")
```

With this annotation procedure, reliability already has improved a lot.

### Choosing few-shot examples

A few general rules for few-shot examples. You want:
- approximately equal distribution of examples for each category (even if they do not occur at an equal rate in your data)
- at least one general, easy case for each label
- additionally, exemplary cases for your decision rules. If you give explanations, it can help to explicate the rule you used to decide the case to reinforce decision rules

DO NOT take examples from your data as few-shot examples - this might bias results, especially if you re-use those data within your validation steps. Instead, write new examples - or let an llm write them (but be careful to curate them to: make them sound natural, not overdo markers and key phrases. Also, preferable do not use the same llm for few shot example generation and annotation)

Once you have a set of cases to choose from, another utility function can come useful: liknites runs over your few-shot examples, leaving out all labels except one and then tests if it would label this one case correctly. It gives you agreement to show which label would be labeled correctly the most if left out. This method is called leave-one-out testing. (Liknites is another name for the god Bacchus, it means: "he of the winnowing fan" - as wikipedia notes "A winnowing fan was used to separate the chaff from the grain.")

Generally, you will want: Only a few (often: one) clear case per label that would achieve high agreement - those are there to keep you grounded to the core concept. The rest should be hard cases - those it gets wrong more often if the example is not in the list. Curate your few-shot examples for a good balance of those cases.

```{r example3_2_4, eval = FALSE}
library(bacchuss)

few_shot_curration <- liknites(n_runs = 3,
                             instructions = example_instructions_expl,
                             examples = example_fewshot$input,
                             explanations = example_fewshot$explanation,
                             codes = example_fewshot$label,
                             reminder = example_reminder_expl,
                             model = "mistral-nemo")

View(few_shot_curration)
```

### Where to put few-shot examples

In standard use, bacchuss expects you to hand over examples, codes and explanations and then puts them in the user-messages and assistant responses sent to the llm. Putting examples into the system prompt can achieve similar validity. If you want to do that instead, add them to your system prompt and leave those variables NULL for bacchuss (or TRUE for explanations, if applicable).

## Step 3.3: Hard cases with few-shot instructions

You can also test the degree of insecurity left over with your new instructions - briseus also works with few-shot and chain of thought. That means you can identify which of the text examples in your sample set generate disagreement in multiple runs

```{r example3_3, eval = FALSE}
library(bacchuss)

few_cot_test_3_3 <- briseus(example_set, input_column = "paragraphs", n_runs = 5,
                            temperature = 0.7,
                            input_column = "paragraphs",
                            instructions = example_instructions_expl,
                            examples = example_fewshot$input,
                            explanations = example_fewshot$explanation,
                            codes = example_fewshot$label,
                            reminder = example_reminder_expl,
                            model = "mistral-nemo",
                            host = NULL,
                            api_key = NULL)

View(few_cot_test_3_3)
```


## Step 3.4: Compare different llms

Finally, comparing different LLMs can be useful - some are better at your annotation task than others.

Generally, try to write good annotation instructions first and do not adapt too much to peculiarities of one llm - instead, write them agnostically and then choose what llm works best with them.


## Repeat steps 3.1-3.4 until you achieve acceptable results

## Step 4: Representative human validation

Once the annotation of the non-representative sample is acceptable, you should test if the annotation works with a representative, real-world sample.
Let your LLM annotate a large enough sample for human validation. I currently do not have data for these steps in the package data set.

```{r example4, eval = FALSE}
library(bacchuss)

example_few_cot <- bacchuss(your_data, input_column = "paragraphs",
                            instructions = example_instructions_expl,
                            examples = example_fewshot$input,
                            explanations = example_fewshot$explanation,
                            codes = example_fewshot$label,
                            reminder = example_reminder_expl,
                            model = "mistral-nemo",
                            host = NULL,
                            api_key = NULL)

View(example_few_cot)
```

## Step 4.1: Unweighed or weighed human annotation sampling

Draw a sample from the LLM-annotated set. Depending on how prevalent your dimension of interest is choose random sampling or weighed sampling. Weighed sampling means oversampling the (rarer) positive category and adjusting validation measurements to that weighing. I will later ad a vignette that explains this step.

## Step 4.2: Measuring human reliability and LLM validity

Export a sample without llm labels and instruct human annotators to label your data. They can use the same annotation instructions as your llm. 
As with regular human annotation, plan for annotator training:
- before letting them annotate the real sample, let them code smaller training samples and discuss unclear cases
- spend enough time on explaining dimensions, looking over examples, discussing mismatching annotations before annotating the real sample
- Only once you are happy with their performance do the real human gold standard data annotation step

For an overview over human annotation I again suggest the literature on human annotation - for example, Kimberly Neuendorf's "The Content Analysis Guidebook".

One note: If your human annotators participated in steps 2 - 3, this very likely decreases training times necessary for the human annotation step.

## Step 4.3: Final analysis or further improvements

Validate how well your human annotations match the LLM's labels. If you are happy with the results, you can move on to step 5.

## Step 4.3.1: If necessary, improve annotation instructions, remember accurate weighing

If you are unhappy with your results, you can improve annotation instructions in a last round to see if you can improve results. If you do that, you need to report it with the results, and, depending on the scale of your changes, might need to test validity with another human annotated validation set. 

## Step 5: Report results

If you followed the steps of this procedure, you should report these steps in your final analysis. Most papers I have seen thus far often report "we improved the prompt in an iterative process" with sparse details, so if you report more details than that, reviewers will be very grateful. At the same time, many details of this process probably should go into the Appendix - just describe steps 1, 2, and 3 in a short paragraph, focus your reporting on step 4, and put details into the Appendix, next to your final annotation instructions. For future research it is also very useful if you put your procedure on an online appendix, github or osf. Your colleagues will thank you.